### 2025-07-17 paper/sections/intro.qmd
major adjustments to intro and related work

---
bibliography: bibliography.bib
---

@@

Counterfactual explanations (CE) have become an integral part of the broader explainable AI (XAI) toolkit [@molnar2022interpretable]. Originally proposed by @wachter2017counterfactual, CEs prescribe minimal changes for factual inputs that, if implemented, would prompt some fitted model to produce an alternative, more desirable output. This is useful and necessary to not only understand how opaque AI models make predictions, but also to provide recourse to individuals subjected to them: a retail bank, for example, could use CE to provide meaningful feedback to unsuccessful loan applicants that were rejected based on an opaque AI decision-making system (ADM).

@@

Existing state-of-the-art (SOTA) approaches in the field have largely focused on designing model-agnostic CE methods that identify subsets of counterfactuals, which comply with specific desiderata. This is problematic, because a narrow focus on any specific desideratum can adversely affect others: it is possible, for example, to generate plausible counterfactuals for models that are also highly vulnerable to implausible, possibly adversarial counterfactuals [@altmeyer2024faithful]. In this work, we therefore follow the paradigm that models---as opposed to explanation methods---should be held accountable for plausible and actionable explanations. While previous work has shown that at least plausibility can be indirectly achieved through existing techniques aimed at models' generative capacity, generalization and robustness [@altmeyer2024faithful; @augustin2020adversarial; @schut2021generating], we directly incorporate both plausibility and actionability in the training objective of models.

Specifically, **we propose Counterfactual Training (CT)**: a novel training regime for differentiable models that leverage counterfactual explanations on-the-fly to ensure that models learn plausible and actionable explanations for the data. In essence, we achieve this by contrasting faithful explanations with ground-truth data, while at the same time protecting features that are subject to actionability constraints. To the best of our knowledge, CT represents the first attempt in this direction.

In our experiments, we show that thanks to counterfactual training, (1) implausibility decreases with 16% on average; (2) the cost of reaching counterfactuals with protected features decreases with 15%; and (3) robust accuracy increases for all datasets with up to .42.

__________________________________________________________

The remainder of this manuscript is structured as follows. @sec-lit presents related work, focusing on the link between AEs and CEs. Then follow our two principal contributions. In @sec-method, we introduce our methodological framework and show theoretically that it can be employed to enforce global actionability constraints. In @sec-experiments, through extensive experiments, we empirically demonstrate that CT substantially improves explainability and positively contributes to the robustness of trained models without sacrificing predictive performance. Finally, in @sec-discussion, we discuss open challenges and conclude that CT is a promising approach towards making opaque models more trustworthy.

>>>>

@@

Today’s prominence of artificial intelligence (AI) has largely been driven by the success of representation learning with large degrees of freedom: instead of relying on features and rules hand-crafted by humans, modern machine learning (ML) models are tasked with learning highly complex representations directly from the data, guided by narrow objectives such as predictive accuracy [@goodfellow2016deep]. These models tend to be so complex and opaque that humans cannot easily interpret their decision-making.

Counterfactual explanations (CE) have become an integral part of the broader explainable AI (XAI) toolkit [@molnar2022interpretable] that can be employed to make sense of this complexity. Originally proposed by @wachter2017counterfactual, CEs prescribe minimal changes for factual inputs that, if implemented, would prompt some fitted model to produce an alternative, more desirable output (@fig-poc). This is useful and necessary to not only understand how opaque models make predictions, but also to provide recourse to individuals subjected to them: a retail bank, for example, could use CE to provide meaningful feedback to unsuccessful loan applicants that were rejected based on an opaque AI decision-making system (ADM).

@@

Existing state-of-the-art (SOTA) approaches in the field have largely focused on designing model-agnostic CE methods that identify subsets of counterfactuals, which comply with specific desiderata. This is problematic, because a narrow focus on any specific desideratum can adversely affect others: it is possible, for example, to generate plausible counterfactuals for models that are also highly vulnerable to implausible, possibly adversarial counterfactuals [@altmeyer2024faithful]. In this work, we therefore follow the paradigm that models---as opposed to explanation methods---should be held accountable for plausible and actionable explanations. While previous work has shown that at least plausibility can be indirectly achieved through existing techniques aimed at models' generative capacity, generalization and robustness [@altmeyer2024faithful; @augustin2020adversarial; @schut2021generating], we directly incorporate both plausibility and actionability in the training objective of models to improve their overall explanatory capacity.

Specifically, **we propose Counterfactual Training (CT)**: a novel training regime that leverages counterfactual explanations on-the-fly to ensure that differentiable models learn plausible and actionable explanations for the underlying data, while at the same time also being more robust to adversarial examples (AE). @fig-poc illustrates the outcome of CT compared to conventionally trained models: in panel (a), faithful and valid counterfactuals end up near the decision boundary forming a clearly distinguishable cluster in the target class (orange); in panel (b), CT is applied to the same underlying linear classifier architecture resulting in much more plausible counterfactuals; in panel (c), the classifier is again trained conventionally and we have introduced a mutability constraint on the *age* feature at test time---counterfactuals are valid but the classifier is roughly equally sensitive to both features; by contrast, the decision boundary in panel (c) has titled, making the model trained with CT relatively less sensitive to the immutable *age* feature. To achieve these outcomes, CT draws inspiration from the literature on contrastive and robust learning: we contrast faithful CE with ground-truth data while protecting immutable features, and capitalize on methodological links between CE and AE by penalizing the model's adversarial loss on interim (*nascent*) counterfactuals. To the best of our knowledge, CT represents the first venture in this direction with promising results, both empirical and theoretical.

The remainder of this manuscript is structured as follows. @sec-lit presents related work, focusing on the links to contrastive and robust learning. Then follow our two principal contributions. In @sec-method, we introduce our methodological framework and show theoretically that it can be employed to enforce global actionability constraints. In our experiments (@sec-experiments), we find that thanks to counterfactual training, (1) implausibility of CE decreases by 16% on average; (2) the cost of reaching valid counterfactuals with protected features decreases by 15%; and (3) robust accuracy increases for all datasets by up to 42 percentage points. Finally, in @sec-discussion, we discuss open challenges and conclude.

![Counterfactual explanations (stars) for linear classifiers trained under different regimes on synthetic data: (a) conventional training, all mutable; (b) CT, all mutable; (c) conventional, *age* immutable; (d) CT, *age* immutable. The linear decision boundary is shown in green along with training data colored according to ground-truth labels: $y^-=1$ (blue) and $y^+=2$ (orange). Stars indicate counterfactuals in the target class. Synthetic features are annotated as *debt* and *age* for illustrative purposes.](/paper/figures/poc.svg){#fig-poc fig-env="figure*"}

### 2025-07-17 paper/sections/related.qmd
major adjustments to intro and related work

To make the desiderata for our framework more concrete, we follow @augustin2020adversarial in tying explainability to the quality of CEs that can be generated for a given model. The authors show that CEs (understood as minimal input perturbations that yield some desired model prediction) tend to be more meaningful if the underlying model is more robust to adversarial examples. We can make intuitive sense of this finding if we look at adversarial training (AT) through the lens of representation learning with high degrees of freedom. As argued before, learned representations may be sensitive to producing implausible explanations and mispredicting for worst-case counterfactuals (i.e., AEs). Thus, by inducing models to "unlearn" susceptiblity to such examples, adversarial training can effectively remove implausible explanations from the solution space.

## Adversarial Examples are Counterfactuals

The interpretation of the link between explainability through counterfactuals on the one side, and robustness to adversarial examples on the other is backed by empirical evidence. @sauer2021counterfactual demonstrate that using counterfactual images during classifier training improves model robustness. Similarly, @abbasnejad2020counterfactual argue that counterfactuals represent potentially useful training data in machine learning, especially in supervised settings where inputs may be reasonably mapped to multiple outputs. They, too, show that augmenting the training data of (image) classifiers can improve generalization performance. Finally, @teney2020learning argue that counterfactual pairs tend to exist in training data. Hence, their approach aims to identify similar input samples with different annotations and ensure that the gradient of the classifier aligns with the vector between such pairs of counterfactual inputs using a cosine distance loss function.

@@

## Aligning Representations with Explanations

Improving the adversarial robustness of models is not the only path towards aligning representations with plausible explanations. In a closely related work, @altmeyer2024faithful show that explainability can be improved through model averaging and refined model objectives. They propose a way to generate counterfactuals that are maximally faithful to the model in that they are consistent with what the model has learned about the underlying data. Formally, they rely on tools from energy-based modelling [@teh2003energy] to minimize the divergence between the distribution of counterfactuals and the conditional posterior over inputs learned by the model. Their counterfactual explainer, *ECCCo*, yields plausible explanations if and only if the underlying model has learned representations that align with them. The authors find that both deep ensembles [@lakshminarayanan2016simple] and joint energy-based models (JEMs) [@grathwohl2020your] tend to do well in this regard.

Once again it helps to look at these findings through the lens of representation learning with high degrees of freedom. Deep ensembles are approximate Bayesian model averages, which are particularly effective when models are underspecified by the available data [@wilson2020case]. Averaging across solutions mitigates the aforementioned risk of overrelying on a single locally optimal representation that corresponds to semantically meaningless explanations for the data. Likewise, previous work of @schut2021generating found that generating plausible ("interpretable") CEs is almost trivial for deep ensembles that have undergone adversarial training. The case for JEMs is even clearer: they optimize a hybrid objective that induces both high predictive performance and strong generative capacity [@grathwohl2020your], which bears resemblance to the idea of aligning models with plausible explanations and has inspired our CT objective.

>>>>

To make the desiderata for our framework more concrete, we follow previous work in tying the explanatory capacity of models to the quality of CEs that can be generated for them [@altmeyer2024faithful; @augustin2020adversarial]. For simplicity, we refer to 'explanatory capacity' simply as 'explainability' in what follows.

## Explainability and Contrastive Learning

In a closely related work, @altmeyer2024faithful show that model averaging and, in particular, contrastive model objectives can produce more explainable and hence trustworthy models. To do so, the authors first propose a way to generate counterfactuals that are maximally faithful to the model in that they are consistent with what the model has learned about the underlying data. Formally, they rely on tools from energy-based modelling [@teh2003energy] to minimize the divergence between the distribution of counterfactuals and the conditional posterior over inputs learned by the model. Their counterfactual explainer, *ECCCo*, yields plausible explanations if and only if the underlying model has learned representations that align with them. The authors find that both deep ensembles [@lakshminarayanan2016simple] and joint energy-based models (JEMs) [@grathwohl2020your]---a form of constrastive learning---tend to do well in this regard.

It helps to look at these findings through the lens of representation learning with high degrees of freedom. Deep ensembles are approximate Bayesian model averages, which are particularly effective when models are underspecified by the available data [@wilson2020case]. Averaging across solutions mitigates the aforementioned risk of overrelying on a single locally optimal representation that corresponds to semantically meaningless explanations for the data. Likewise, previous work of @schut2021generating found that generating plausible ("interpretable") CEs is almost trivial for deep ensembles that have undergone adversarial training. The case for JEMs is even clearer: they optimize a hybrid objective that induces both high predictive performance and strong generative capacity [@grathwohl2020your], which bears resemblance to the idea of aligning models with plausible explanations and has inspired our CT objective.

## Explainability and Robust Learning

@augustin2020adversarial show that CEs tend to be more meaningful ("explainable") if the underlying model is more robust to adversarial examples. Once again, we can make intuitive sense of this finding if we look at adversarial training (AT) through the lens of representation learning with high degrees of freedom: highly complex and flexible models may learn representations that make them sensitive to implausible or even adversarial examples [@szegedy2013intriguing]. Thus, by inducing models to "unlearn" susceptibility to such examples, adversarial training can effectively remove implausible explanations from the solution space.

This interpretation of the link between explainability through counterfactuals on the one side, and robustness to adversarial examples on the other is backed by empirical evidence. @sauer2021counterfactual demonstrate that using counterfactual images during classifier training improves model robustness. Similarly, @abbasnejad2020counterfactual argue that counterfactuals represent potentially useful training data in machine learning, especially in supervised settings where inputs may be reasonably mapped to multiple outputs. They, too, show that augmenting the training data of (image) classifiers can improve generalization performance. Finally, @teney2020learning argue that counterfactual pairs tend to exist in training data. Hence, their approach aims to identify similar input samples with different annotations and ensure that the gradient of the classifier aligns with the vector between such pairs of counterfactual inputs using a cosine distance loss function.

@@
